---
title: "Contrast analysis"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, estimate, marginal means, emmeans, contrast analysis, post hoc, pairwise t tests]
vignette: >
  %\VignetteIndexEntry{Contrast analysis}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(comment = ">", dpi = 450)
options(digits = 2)

if (!requireNamespace("ggplot2", quietly = TRUE) ||
  !requireNamespace("see", quietly = TRUE) ||
  !requireNamespace("rstanarm", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}

set.seed(333)
```

Warning: we will go **full Bayesian** in this vignette. If you're not familiar
with the Bayesian framework, we recommend starting with [**this gentle
introduction**](https://easystats.github.io/bayestestR/articles/bayestestR.html).

# Testing pairwise differences

In the [**previous
tutorial**](https://easystats.github.io/estimate/articles/estimate_means.html),
we computed marginal at the 3 different `Species` levels from the
[`iris`](https://en.wikipedia.org/wiki/Iris_flower_data_set) dataset. However,
one might also want to **statistically test** the differences between each
levels, which can be achieved through **contrast analysis**. Although the
procedure is much more powerful, its aim is somehow analog to the ***post hoc***
analysis (pretty much consisting of pairwise t-tests) heavily used in
psychological science to palliate the uselessness of ANOVAs.

Let's do that based on the simple model from the previous tutorial:

```{r message=FALSE, warning=FALSE, eval=FALSE}
library(ggplot2)
library(see)
library(rstanarm)
library(modelbased)


model <- stan_glm(Sepal.Width ~ Species, data = iris)
means <- estimate_means(model)
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
library(see)
library(rstanarm)
library(modelbased)

model <- stan_glm(Sepal.Width ~ Species, data = iris, refresh = 0)
means <- estimate_means(model)
```
```{r message=FALSE, warning=FALSE}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_violin() +
  geom_jitter2(width = 0.05, alpha = 0.5) +
  geom_line(data = means, aes(y = Mean, group = 1), size = 1) +
  geom_pointrange(data = means, aes(y = Mean, ymin = CI_low, ymax = CI_high), size = 1, color = "white") +
  theme_modern()
```

**Contrast analysis** can be achieved through the `estimate_contrasts`
function:

```{r message=FALSE, warning=FALSE}
estimate_contrasts(model)
```

As we can see here, all pairwise differences can be considered as significant.

# Complex model

Again, as contrast analysis is based on marginal means, it can be applied to
more complex models:

```{r message=FALSE, warning=FALSE, eval=FALSE}
model <- stan_glm(Sepal.Width ~ Species * Petal.Width, data = iris)
estimate_contrasts(model)
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
model <- stan_glm(Sepal.Width ~ Species * Petal.Width, data = iris, refresh = 0, iter = 1000, chains = 2)
estimate_contrasts(model)
```

For instance, if we add `Petal.Width` in the model, we can see that the
difference between *versicolor* and *virginica* becomes not significant (and
even changes sign).

Note that we can plot simple contrast analysis through **lighthouse plots** with
the help of the `see` package:

```{r message=FALSE, warning=FALSE}
library(see)

plot(estimate_contrasts(model), estimate_means(model)) +
  theme_modern()
```

These represent the estimated means and their CI range (in black), while the
gray areas show the CI range of the difference (as compared to the point
estimate).

# Changes in difference

Interestingly, we can also see how these differences are modulated by another
continuous variable. Based on the model above (including the interaction with
`Petal.Width`), we will compute the contrasts at 100 equally-spaced points of
`Petal.Width`, that we will then visualise.

```{r message=FALSE, warning=FALSE, fig}
contrasts <- estimate_contrasts(model, modulate = "Petal.Width", length = 100)

# Create a variable with the two levels concatenated
contrasts$Contrast <- paste(contrasts$Level1, "-", contrasts$Level2)

# Visualise the changes in the differences
ggplot(contrasts, aes(x = Petal.Width, y = Difference)) +
  geom_ribbon(aes(fill = Contrast, ymin = CI_low, ymax = CI_high), alpha = 0.2) +
  geom_line(aes(colour = Contrast), size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_modern() +
  ylab("Difference")
```

As we can see, the difference between *versicolor* and *virginica* increases as
`Petal.Width` increases. In conclusion, contrast analysis is a powerful tool to
interpret and understand statistical models.

# References

