---
title: "Visualize effects and interactions"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, estimate, estimate link, predictions]
vignette: >
  %\VignetteIndexEntry{Visualize effects and interactions}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(comment = ">", dpi = 450)
options(digits = 2)

if (!requireNamespace("ggplot2", quietly = TRUE) ||
  !requireNamespace("see", quietly = TRUE) ||
  !requireNamespace("gganimate", quietly = TRUE) ||
  !requireNamespace("rstanarm", quietly = TRUE) ||
  !requireNamespace("dplyr", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}

set.seed(333)
```

This vignette will present how to visualize the effects and interactions using
`estimate_link()`.

# Simple regression

## Linear relationship

```{r message=FALSE, warning=FALSE, eval=FALSE}
library(modelbased)

model <- glm(Sepal.Length ~ Sepal.Width, data = iris)

visualization_data <- modelbased::estimate_link(model)
head(visualization_data)
```

<!-- Demo using 'see' instead? -->

```{r message=FALSE, warning=FALSE, eval=FALSE}
library(dplyr)
library(ggplot2)
library(see)

visualization_data %>%
  ggplot(aes(x = Sepal.Width, y = Predicted)) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high), alpha = 0.2) +
  geom_line() +
  see::theme_modern()
```

# More complex regressions

## Non-linear relationships

Note that non-linear relationships can be also described by linear
approximations using
[**estimate_smooth**](https://easystats.github.io/modelbased/articles/estimate_smooth.html).

### Polynomial

```{r message=FALSE, warning=FALSE, eval=FALSE}
glm(Sepal.Length ~ poly(Sepal.Width, 2), data = iris) %>%
  modelbased::estimate_link() %>%
  ggplot(aes(x = Sepal.Width, y = Predicted)) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high), alpha = 0.2) +
  geom_line() +
  see::theme_modern()
```

### Additive Models

```{r message=FALSE, warning=FALSE, eval=FALSE}
library(mgcv)

mgcv::gam(Sepal.Length ~ s(Sepal.Width), data = iris) %>%
  modelbased::estimate_link() %>%
  ggplot(aes(x = Sepal.Width, y = Predicted)) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high), alpha = 0.2) +
  geom_line() +
  see::theme_modern()
```

# Interactions

## Interaction with a factor

TODO.

## Interaction with another continuous variable

TODO.

# Supported Models

TODO.

# References

