---
title: "What are, why use and how to get marginal means"
output: 
  github_document:
    toc: true
    fig_width: 10.08
    fig_height: 6
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, estimate, marginal means, emmeans]
vignette: >
  %\VignetteIndexEntry{Marginal means}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
knitr::opts_chunk$set(comment = ">", dpi = 450)
options(digits = 2)

if (!requireNamespace("ggplot2", quietly = TRUE) ||
    !requireNamespace("see", quietly = TRUE) ||
    !requireNamespace("rstanarm", quietly = TRUE) ||
    !requireNamespace("dplyr", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}

set.seed(333)
```


This vignette will introduce the concept of marginal means. Warning: we will go **full Bayesian**. If you're not familiar with the Bayesian framework, we recommend starting with [**this gentle introduction**](https://easystats.github.io/bayestestR/articles/bayestestR.html).

# Raw Means


The [`iris`](https://en.wikipedia.org/wiki/Iris_flower_data_set) dataset, available in base R, contains observations of 3 types of iris flowers (the `Species` variable); *Setosa*, *Versicolor* and *Virginica*, for which different features were measured, such as the length and width of the sepals and petals.

A traditional starting point, when reporting such data, is to start by descriptive statistics. For instance, **what is the mean `Sepal.Width` for each of the 3 species**.

We can compute the means very easily by grouping the observations by species, and then computing the mean and the SD:

```{r message=FALSE, warning=FALSE}
library(dplyr)

iris %>% 
  group_by(Species) %>% 
  summarise(Mean_Sepal.Width = mean(Sepal.Width),
            SD_Sepal.Width = sd(Sepal.Width))
```

We can also provide a plot:

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(see)

ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_violin() +
  geom_jitter2(width = 0.05) +
  theme_modern()
```


However, these **raw means** might be biased, as the number of observations in each group might be different. Moreover, there might some hidden covariance or mediation with other variables in the dataset, creating a "spurious" influence on the means. **How to take these things into account?**

# Marginal Means

Another way of analyzing the means is to actually statistically **model them**, rather than simply describe them as they appear in the data. For instance, we could fit a simple Bayesian linear regression modeling the relationship between `Species` and `Sepal.Width`.

Marginal means are basically means extracted from a statistical model. Note that as we are in a Bayesian framework, we will report the **median** of the posterior distribution of the marginal means.

```{r message=FALSE, warning=FALSE, eval=FALSE}
library(rstanarm)
library(modelbased)

model <- stan_glm(Sepal.Width ~ Species, data = iris)
means <- estimate_means(model)
means
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
library(rstanarm)
library(modelbased)

model <- stan_glm(Sepal.Width ~ Species, data = iris, refresh = 0)
means <- estimate_means(model)
means
```


We can now add these means, as well as the [**credible interval (CI)**](https://easystats.github.io/bayestestR/articles/credible_interval.html) representing the uncertainty of the estimation, as an overlay on the previous plot:

```{r message=FALSE, warning=FALSE}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_violin() +
  geom_jitter2(width = 0.05, alpha = 0.5) +
  
  geom_line(data = means, aes(y = Mean, group = 1), size = 1) +
  geom_pointrange(data = means, aes(y = Mean, ymin = CI_low, ymax = CI_high), size = 1, color = "white") +
  theme_modern()
```


# Complex Models

The power of marginal means resides in the fact that they can be estimated from much more complex models. For instance, we could fit a model that takes into account the interaction with the other variables, `Petal.Length` and  `Petal.Width`. The estimated means will be "adjusted" (or will take into account) for variations of these other components.

```{r message=FALSE, warning=FALSE, eval=FALSE}
model <- stan_glm(Sepal.Width ~ Species * Sepal.Length * Petal.Width, data = iris)
means_complex <- estimate_means(model)
means_complex
```
```{r message=FALSE, warning=FALSE, echo=FALSE}
model <- stan_glm(Sepal.Width ~ Species * Sepal.Length * Petal.Width, data = iris, refresh = 0, iter = 1000, chains = 2)
means_complex <- estimate_means(model)
means_complex
```


```{r message=FALSE, warning=FALSE}
ggplot(iris, aes(x = Species, y = Sepal.Width, fill = Species)) +
  geom_violin() +
  geom_jitter2(width = 0.05, alpha = 0.5) +
  
  geom_line(data = means, aes(y = Mean, group = 1), size = 1, alpha = 0.25) +
  geom_pointrange(data = means, aes(y = Mean, ymin = CI_low, ymax = CI_high), size = 1, color = "white") +
  
  geom_line(data = means_complex, aes(y = Mean, group = 1), size = 1) +
  geom_pointrange(data = means_complex, aes(y = Mean, ymin = CI_low, ymax = CI_high), size = 1, color = "yellow") +
  theme_modern()
```

That's interesting: it seems that when adjusting the model for petal characteristics, the differences between Species seems to be even bigger!

**But are these differences "significant"?** Click [here to read the tutorial on **contrast analysis**](https://easystats.github.io/estimate/articles/contrast_analysis.html).


# References
