---
title: "Visualize effects and interactions"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, estimate, estimate link, estimate relation, predictions]
vignette: >
  %\VignetteIndexEntry{Visualize effects and interactions}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(comment = ">", dpi = 300)
options(digits = 2)

if (!requireNamespace("ggplot2", quietly = TRUE) ||
  !requireNamespace("mgcv", quietly = TRUE) ||
  !requireNamespace("see", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}

set.seed(333)
```

This vignette will present how to visualize the effects and interactions using
`estimate_relation()`.

Note that the statistically correct name of `estimate_relation` is `estimate_expectation` (which can be used as an alias), as it refers to expected predictions (read [more](https://easystats.github.io/insight/reference/get_predicted.html)).

## Simple regression

### Linear relationship

```{r}
library(modelbased)

model <- lm(Sepal.Length ~ Sepal.Width, data = iris)

visualization_data <- estimate_relation(model)
head(visualization_data)
```


```{r}
library(ggplot2)
plot(visualization_data, line = list(color = "red")) +
  theme_minimal()
```

## More complex regressions

### Polynomial

```{r}
lm(Sepal.Length ~ poly(Sepal.Width, 2), data = iris) |>
  modelbased::estimate_relation(length = 50) |>
  plot()
```

### Additive Models

```{r}
library(mgcv)

mgcv::gam(Sepal.Length ~ s(Sepal.Width), data = iris) |>
  modelbased::estimate_relation(length = 50) |>
  plot()
```


## References
