---
title: "Mixed effects models"
output:
  rmarkdown::html_vignette:
    toc: true
tags: [r, programing]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Mixed effects models}
  %\VignetteEngine{knitr::rmarkdown}
bibliography: bibliography.bib
---

```{r set-options, echo = FALSE, message = FALSE}
library(knitr)
knitr::opts_chunk$set(
  package.startup.message = FALSE,
  dpi = 300,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = TRUE
)
options(
  knitr.kable.NA = "",
  digits = 2,
  width = 800,
  modelbased_join_dots = FALSE,
  modelbased_select = "minimal"
)
pkgs <- c("lme4", "marginaleffects", "collapse", "Formula")
successfully_loaded <- suppressPackageStartupMessages(vapply(pkgs, requireNamespace, FUN.VALUE = logical(1L), quietly = TRUE))
can_evaluate <- all(successfully_loaded)

if (can_evaluate) {
  knitr::opts_chunk$set(eval = TRUE)
  junk <- suppressMessages(suppressPackageStartupMessages(vapply(pkgs, require, FUN.VALUE = logical(1L), quietly = TRUE, character.only = TRUE)))
} else {
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{css, echo=FALSE, eval = TRUE}
.custom_note {
  border-left: solid 5px hsl(220, 100%, 30%);
  background-color: hsl(220, 100%, 95%);
  padding: 10px;
  margin-bottom: 10px;
  font-size: 0.9em;
}
```

## Estimated Marginal Means in Mixed Effects Models: Navigating Conditional and Marginal Effects

Mixed models, with their ability to account for hierarchical or clustered data, offer powerful tools for understanding complex relationships. A key aspect of interpreting these models is understanding **estimated marginal means (EMMs)**, which represent the predicted outcome for specific groups or conditions, while holding other variables constant. However, calculating EMMs in mixed models is not always straightforward, and the results can vary depending on the approach taken.

One crucial distinction is between **conditional** and **marginal** predictions (or effects). Conditional predictions are specific to a particular level of the random effect (e.g., the predicted outcome for a specific individual in a study). Marginal predictions, on the other hand, average over the random effects, providing an overall estimate of the effect in the population. This is a crucial difference, as the marginal effect is often the quantity of interest when we want to generalize to the population [@heiss_marginal_2022].

::: custom_note
<p>Based on the definitions from Heiss (2022), we can say that the conditional vs. marginal distinction applies to any sort of hierarchical structure (<em>clusters</em>) in multilevel models:</p>

- **Conditional effect** = the effect of a variable in an *average cluster* (i.e., group-specific, subject-specific or cluster-specific effect, or an average or a typical cluster, `re.form = NA`)
- **Marginal effect** = effect of a variable *across clusters on average* (i.e., global/population-level effect, or clusters on average, `re.form = NULL`)

For `backend = "marginaleffects"`, the `re.form` argument is set to `NULL` for mixed models by default, to calculate _marginal predictions_. You can use for instance `re.form = NA` in your `estimate_means()` call to change the default value (`NA` will produce _conditional predictions_).

As outlined in the [vignette on marginalization methods](https://easystats.github.io/modelbased/articles/technical_marginalization.html), marginalization in mixed models involves two distinct dimensions: marginalizing across *random effects (clusters)* and marginalizing across *covariates (observations)*.

We can produce *marginal predictions* regarding the higher-level units (random effects) *conditioned on* a "typical" observation based on a balanced data grid, or generate predictions that are *fully marginalized* across both clusters and observations.

The default (`estimate = "typical"`) calculates predictions for a balanced data grid representing all combinations of focal predictor levels (specified in `by`), which is particularly useful for comparing groups. Setting `estimate = "average"` or `estimate = "population"` is useful when you want to calculate the average expected outcome across the actual observations from the sample at hand.

Note that for *linear* mixed models, `re.form = NA` will only have an effect when `estimate = "average"` or `estimate = "population"`.
:::

## Linear mixed models

### Balanced Data

This section demonstrates the calculation of estimated marginal means (EMMs) in a linear mixed model using balanced data. We'll use the `sleepstudy` dataset from the **lme4** package.

In this example, we fit a linear mixed model predicting `Reaction` based on `Days`, with random intercepts and slopes for `Subject`. Because the data is balanced and we have a linear model, results are identical for all marginalization methods.

```{r}
library(modelbased)
data(sleepstudy, package = "lme4")
model <- lme4::lmer(Reaction ~ Days + (1 + Days | Subject), data = sleepstudy)

# `estimate = "typical"`, `re.form = NULL`: a "typical" observation
# across clusters on average
estimate_means(model, "Days")

# `estimate = "typical"`, `re.form = NA`: a "typical" observation
# for a "typical" (average) cluster - `re.form = NA` has no effect here
estimate_means(model, "Days", re.form = NA)

# fully marginalized predictions, still same result as above
estimate_means(model, "Days", estimate = "average")
```

### Imbalanced Data

This section explores the impact of imbalanced data on EMM calculations in linear mixed models. We'll use the `penguins` dataset, which has imbalanced groups that we use as higher-level unit, as well as imbalanced predictors.

Since `estimate = "typical"` creates a balanced data grid, *including* the levels of the random effects, we get a "typical" observation marginalized over a *balanced* grid of random effects levels. In this case, setting `re.form = NA` has no effect when we have *linear* mixed models.

```{r}
data(penguins)
# Add a non-focal categorical covariate to illustrate differences
# in marginalization
penguins$body_mass <- as.factor(datawizard::categorize(penguins$body_mass))
model <- lme4::lmer(bill_len ~ sex + body_mass + (1 | species), data = penguins)

# conditional predictions marginalized over
# a "balanced" grid of random effects
estimate_means(model, "sex")

# (fully) marginal predictions
estimate_means(model, "sex", estimate = "average")

# marginal predictions conditioning on a "typical" cluster
estimate_means(model, "sex", estimate = "average", re.form = NA)
```

### Excourse: Technical Details

::: custom_note
To better understand how these predictions are computed under the hood, let's break down the underlying workflow. The code below illustrates the manual step-by-step procedures for grid-based versus sample-averaged predictions.

Technically, `estimate_means(model, "sex")` is equivalent to

1. creating a balanced data grid of all combinations of predictor levels including random effect levels
2. predicting the outcome for the artificial observations represented by that data grid
3. followed by averaging the predictions by the levels of the focal predictors.

```{r}
dg <- insight::get_datagrid(model, "sex", factors = "all", include_random = TRUE)
out <- cbind(predicted = predict(model, newdata = dg), dg)
aggregate(out$predicted, list(sex = out$sex), mean)
```

`estimate_means(model, "sex", estimate = "average")` is equivalent to

1. predicting the outcome for each observation in the data
2. followed by averaging the predictions by the levels of the focal predictors.

```{r}
out <- cbind(predicted = predict(model), insight::get_data(model))
aggregate(out$predicted, list(sex = out$sex), mean)
```
:::

## Generalized linear mixed models

The generalized linear mixed model (GLMM) example, using a Poisson distribution to model fish counts, demonstrates the substantial differences between marginal and conditional predictions in non-linear models. When examining the effect of the `camper` variable, marginal predictions yield higher estimated means (1.26 for `camper = 0` and 3.21 for `camper = 1`) compared to conditional predictions (using `re.form = NA`), which are lower (0.66 for `camper = 0` and 1.68 for `camper = 1`).

```{r}
data("fish", package = "insight")
model <- lme4::glmer(
  count ~ child + camper + (1 | persons),
  data = fish,
  family = poisson()
)

# conditional predictions marginalized over
# a "balanced" grid of random effects
estimate_means(model, "camper")

# conditional predictions for a "typical" cluster
estimate_means(model, "camper", re.form = NA)
```

When using `estimate = "average"` to average across all observations and clusters, the differences between marginal and conditional predictions (regarding random effects) persist, with marginal predictions showing higher means (1.52 and 4.54) than conditional predictions (1.20 and 3.19), but both results differ from the predictions for "typical" observations based on the default `estimate` option, as shown in the previous output.

```{r}
# (fully) marginal predictions
estimate_means(model, "camper", estimate = "average")

# marginal predictions, conditioning on a "typical" cluster
estimate_means(model, "camper", estimate = "average", re.form = NA)
```

This highlights that in GLMMs, the choice between marginal and conditional predictions significantly impacts the estimated means and, consequently, the interpretation of the model's results.

## Conclusion

This vignette has demonstrated the nuances of calculating estimated marginal means (EMMs) in mixed models, highlighting the critical distinction between conditional and marginal predictions. While conditional predictions focus on specific levels of random effects, marginal predictions average over them, providing a population-level perspective. We've seen that in linear mixed models with balanced data, these two approaches often yield similar results. However, in generalized linear mixed models (GLMMs), the differences between conditional and marginal predictions can be substantial.

Furthermore, the default approach of calculating predictions for a balanced data grid, while useful for comparing groups, does not necessarily reflect the actual distribution of observations in the sample. For imbalanced data, this can yield significantly different results even in linear mixed models in comparison to average predictions (using `estimate = "average"`).

**Therefore, we recommend combining marginal predictions (using the default `backend = "marginaleffects"`) with averaging across all observations (by setting `estimate = "average"` or `estimate = "population"`) when working with mixed models.** This approach effectively incorporates the variation inherent in the random effects (higher-level units) and provides EMMs that more accurately reflect the overall patterns observed within the actual sample data. By averaging over the observed data, we obtain a more robust and representative estimate of the population-level effects, making it a preferred strategy for interpreting mixed model results.

```{r echo=FALSE}
# reset options
options(
  modelbased_join_dots = NULL,
  modelbased_estimate = NULL,
  modelbased_select = NULL
)
```

# References
