---
title: "Context Effects: How the Environment Shapes the Individual"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Context Effects: How the Environment Shapes the Individual}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r , include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  out.width = "100%"
)

pkgs <- c(
  "datawizard",
  "parameters",
  "lme4"
)

if (!all(sapply(pkgs, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}

set.seed(333)
```

When analyzing longitudinal or clustered data, a standard mixed-effects model often blends two completely different processes into a single coefficient: the overall differences between individuals (or groups) and the acute fluctuations within an individual. However, separating these effects is just the first step. By evaluating them side-by-side, we can uncover a third, crucial piece of the puzzle: the *context effect*.

The context effect describes the additional influence that a broader environment or a chronic baseline trait has on an individual, independent of their immediate, momentary state [@bell_understanding_2018, @rohrer_these_2023]. By using the `demean()` function to separate our time-varying predictors, we can isolate the within- and between-components, evaluate them independently, and formally test how the overarching context shapes individual outcomes.

To understand the underlying problem of heterogeneity bias, demeaning (person-mean centering), and within/between effects in more detail, it is highly recommended to [read this vignette first](https://easystats.github.io/parameters/articles/demean.html).

# Sample data used in this vignette

```{r}
library(parameters)
data("qol_cancer")
```

* Variables:
  * `QoL` : Response (quality of life of patient)
  * `phq4` : Patient Health Questionnaire, **time-varying** variable
  * `education`: Educational level, **time-invariant** variable, co-variate
  * `ID` : patient ID
  * `time` : time-point of measurement

## Computing the de-meaned and group-meaned variables

To calculate the within- and between-effects, we perform a special way of centering variables called [*demeaning*](https://easystats.github.io/datawizard/reference/demean.html). This "separates" the within-effect from a between-effect of a predictor.

```{r}
qol_cancer <- datawizard::demean(qol_cancer, select = c("phq4", "QoL"), by = "ID")
```

Now we have:

  * `phq4_between`: time-varying variable with the mean of `phq4` across all
    time-points, for each patient (ID).

  * `phq4_within`: the de-meaned time-varying variable `phq4`.

# Calculating the within- and between effects using mixed models

First, we start with calculating the within- and between-effects from `phq4`, before we move on to investigate the context effects.

```{r}
library(lme4)
mixed <- lmer(
  QoL ~ time + phq4_within + phq4_between + education + (1 + time | ID),
  data = qol_cancer
)
# effects = "fixed" will not display random effects, but split the
# fixed effects into its between- and within-effects components.
model_parameters(mixed, effects = "fixed")
```

Looking at the fixed effects output for the `phq4` (Patient Health Questionnaire) variable, we can interpret the coefficients as follows:

- *The Between-Effect (`phq4_between` = -6.13):* This captures the general, trait-like differences across patients. It answers the question: How does a patient's overall average score affect their outcome compared to other patients? If Patient A has an overall average `phq4` score that is 1 unit higher than Patient B's average, we expect Patient A's Quality of Life (QoL) to be 6.13 points lower on average than Patient B's.

- *The Within-Effect (`phq4_within` = -3.72):* This captures the state-like, time-to-time fluctuations for an individual. It answers the question: What happens when a patient deviates from their own baseline? If a patient scores 1 unit higher on the `phq4` at a specific time-point compared to their own personal average, their expected Quality of Life at that specific measurement decreases by an additional 3.72 points.

If we had entered `phq4` into the model as a single, uncentered variable, the resulting coefficient would be a weighted average of these two distinct effects. This can be highly misleading, as it obscures both the overarching patient-to-patient differences and the specific time-to-time dynamics. Separating them provides a much clearer picture of how psychological burden impacts quality of life on multiple levels.

# Context effect - contrasting within- and between-effects

Conceptually, when analyzing clustered or longitudinal data, we are looking at two distinct levels of influence:

- *Individual level (within-effect):* What is the impact of an individual's temporary deviation from their own group mean? This captures state-like fluctuations or acute changes.

- *Group level (between-effect):* What is the impact of the group's general environment or overall average, which affects all members equally? This captures trait-like, baseline differences or overarching environments.

The difference between these two effects is called the *context effect*. A context effect describes the additional influence that the general environment or baseline trait has on an individual, holding their raw, current state constant. It demonstrates that people with identical current, raw values (such as the exact same momentary income or symptom severity) face different outcomes depending on their baseline or the environment in which they live.

To test whether the within- and between-effects are significantly different from each other, we can estimate their contrast:

```{r}
library(modelbased)
estimate_contrasts(mixed, c("phq4_within", "phq4_between"))
```

The output shows a significant contrast of 2.41 between the within- and between-effects. Since the between-effect in our model (-6.13) is stronger (more negative) than the within-effect (-3.72), the context-effect (Between minus Within) is -2.41.

What does this mean practically? Imagine two patients who both report the exact same raw `phq4` score on a given day. However, Patient A generally suffers from a higher overarching psychological burden (their personal `phq4` average is 1 unit higher than Patient B's). The context effect tells us that, despite experiencing the exact same severity of symptoms today, Patient A's expected Quality of Life is an additional 2.41 points lower than Patient B's simply because of their higher baseline burden. The trait-like baseline carries an extra "penalty" for the quality of life that goes beyond mere day-to-day fluctuations.

# Time trends of within- and between-effects

To investigate whether the impact of `phq4` changes over time, we can extend our model by adding interaction terms between the time of measurement (`time`) and our two centered variables (`phq4_within` and `phq4_between`).

```{r}
mixed <- lmer(
  QoL ~ time * (phq4_within + phq4_between) + education + (1 + time | ID),
  data = qol_cancer
)
model_parameters(mixed, effects = "fixed")
```

The results table now shows us whether time has a moderating influence on our effects. To see this, we look at the two interaction terms at the bottom of the table:

- Interaction `time × phq4_within`: The coefficient (0.33) is small and not statistically significant (p = 0.592). This means that the negative effect of individual, temporary fluctuations remains largely stable over time. When a patient's psychological burden acutely increases, their quality of life decreases to a very similar extent at any given time point.

- Interaction `time × phq4_between`: This coefficient (-0.66) is also not statistically significant (p = 0.077). The negative sign suggests that the gap between patients potentially widens slightly over time. The "disadvantage" in quality of life experienced by patients with a generally high psychological baseline burden might further increase over the course of the study compared to less burdened patients.

These temporal dynamics can be best understood visually using Estimated Marginal Means. The generated plots confirm the statistical results from the table very clearly.

```{r}
estimate_means(mixed, c("time", "phq4_within=[sd]")) |> plot()
```

In the first plot (*within-effects over time*), the lines for the different levels of `phq4_within` (mean as well as +/- one standard deviation) run almost parallel. The constant distance between the lines visualizes the lack of interaction: An acute increase in psychological symptoms (shifting to the blue line) depresses the quality of life uniformly across all time points.

```{r}
estimate_means(mixed, c("time", "phq4_between=[sd]")) |> plot()
```

In the second plot (*between-effects over time*), a slight divergence of the lines is visible. While patients with a generally low burden (red line) experience a slight increase in their quality of life over time, the quality of life for patients with a generally high burden (blue line) stagnates at a lower level or even drops minimally.

# Time trends of context effects

Finally, we might ask whether the context effect itself - the difference between the within- and between-effects - is stable, or if the "penalty" of a high baseline burden changes over the course of the study. We can examine this by calculating the marginal contrasts at each specific time point.

```{r}
estimate_contrasts(mixed, c("phq4_within", "phq4_between"), by = "time")
```

The contrast analysis reveals a clear and interesting trajectory: the context effect grows substantially stronger as time progresses.

- *At Time 1 (Baseline):* The difference between the within- and between-effect is relatively small (1.32) and not statistically significant (p = 0.188). This indicates that at the beginning of the observations, it does not matter much whether a patient's psychological burden is acute (a temporary spike) or chronic (a generally high baseline). The immediate impact on their Quality of Life is very similar.

- *At Time 2 and Time 3:* As the study progresses, the contrast becomes highly significant and the gap widens (increasing to 2.31, then 3.29).

**What does this mean practically?**

Over time, having a chronically high baseline of psychological symptoms (the trait-level burden) becomes increasingly detrimental compared to merely experiencing a temporary, acute spike in symptoms. While patients might be able to buffer or cope with an acute worsening of their mental state similarly well at any point, the cumulative "wear and tear" of a chronically high burden takes an increasing toll on their quality of life as time goes on.

# Testing the Change in the Context Effect Over Time

While the previous table showed the context effect at each specific time point, we also need to formally test whether the change in this effect over time is statistically significant. We can do this by computing pairwise comparisons of the context effect across the different time points.

```{r}
estimate_contrasts(mixed, c("phq4_within", "phq4_between", "time"))
```

This pairwise comparison table adds a crucial statistical caveat to our visual and descriptive observations. The Difference column here represents the mathematical change in the size of the context effect between two specific time points (e.g., the context effect grew by 0.99 points from Time 1 to Time 2).

However, looking at the statistics, we can see that none of the differences between the time points are statistically significant (all p-values = 0.183).

**What does this mean practically?**

Although our previous analysis showed the context effect becoming locally significant at Time 2 and Time 3, and descriptively appearing to grow, we do not have enough statistical evidence to claim that the context effect changes over time. The widening gap we observed is not robust enough to rule out random sampling variation.

Therefore, the most accurate conclusion is that the "penalty" of having a chronically high baseline burden (the context effect) is consistently present across the later stages of the study, but its magnitude remains relatively stable rather than strictly worsening over time.

# Disclaimer

Do not base your conclusions solely on the rigid dichotomy of statistical significance.

# References
