---
title: "Context Effects: How the Environment Shapes the Individual"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{Context Effects: How the Environment Shapes the Individual}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  out.width = "100%"
)

pkgs <- c(
  "datawizard",
  "parameters",
  "lme4"
)

if (!all(sapply(pkgs, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}

set.seed(333)
```

When analyzing longitudinal or clustered data, a standard mixed-effects model often blends two completely different processes into a single coefficient: the overall differences between individuals (or groups) and the acute fluctuations within an individual. However, separating these effects is just the first step. By evaluating them side-by-side, we can uncover a third, crucial piece of the puzzle: the *context effect*.

The context effect describes the additional influence that a broader environment or a chronic baseline trait has on an individual, independent of their immediate, momentary state ([@bell_understanding_2018, @rohrer_these_2023]). By using the `demean()` function to separate our time-varying predictors, we can isolate the within- and between-components, evaluate them independently, and formally test how the overarching context shapes individual outcomes.

To understand the underlying problem of heterogeneity bias, demeaning (person-mean centering), and within/between effects in more detail, it is highly recommended to [read this vignette first](https://easystats.github.io/parameters/articles/demean.html).

# Sample data used in this vignette

```{r}
library(parameters)
data("qol_cancer")
```

* Variables:
  * `QoL` : Response (quality of life of patient)
  * `phq4` : Patient Health Questionnaire, **time-varying** variable
  * `education`: Educational level, **time-invariant** variable, co-variate
  * `ID` : patient ID
  * `time` : time-point of measurement

## Computing the de-meaned and group-meaned variables

To calculate the within- and between-effects, we perform a special way of centering variables called [*demeaning*](https://easystats.github.io/datawizard/reference/demean.html). This "separates" the within-effect from a between-effect of a predictor.

```{r}
library(datawizard)
qol_cancer <- demean(qol_cancer, select = c("phq4", "QoL"), by = "ID")
```

Now we have:

  * `phq4_between`: time-varying variable with the mean of `phq4` across all
    time-points, for each patient (ID).

  * `phq4_within`: the de-meaned time-varying variable `phq4`.

# Calculating the within- and between effects using mixed models

First, we start with calculating the within- and between-effects from `phq4`, before we move on to investigate the context effects.

```{r}
library(lme4)
mixed <- lmer(
  QoL ~ time + phq4_within + phq4_between + education + (1 + time | ID),
  data = qol_cancer
)
# effects = "fixed" will not display random effects, but split the
# fixed effects into its between- and within-effects components.
model_parameters(mixed, effects = "fixed") |> display(format = "tt")
```

Looking at the fixed effects output for the `phq4` (Patient Health Questionnaire) variable, we can interpret the coefficients as follows:

- *The Between-Effect (`phq4_between` = -6.13):* This captures the general, trait-like differences across patients. It answers the question: How does a patient's overall average score affect their outcome compared to other patients? If Patient A has an overall average `phq4` score that is 1 unit higher than Patient B's average, we expect Patient A's Quality of Life (QoL) to be 6.13 points lower on average than Patient B's.

- *The Within-Effect (`phq4_within` = -3.72):* This captures the state-like, time-to-time fluctuations for an individual. It answers the question: What happens when a patient deviates from their own baseline? If a patient scores 1 unit higher on the `phq4` at a specific time-point compared to their own personal average, their expected Quality of Life at that specific measurement decreases by an additional 3.72 points.

If we had entered `phq4` into the model as a single, uncentered variable, the resulting coefficient would be a weighted average of these two distinct effects. This can be highly misleading, as it obscures both the overarching patient-to-patient differences and the specific time-to-time dynamics. Separating them provides a much clearer picture of how psychological burden impacts quality of life on multiple levels.

# Context effect - contrasting within- and between-effects

Conceptually, when analyzing clustered or longitudinal data, we are looking at two distinct levels of influence:

- *Individual level (within-effect):* What is the impact of an individual's temporary deviation from their own group mean? This captures state-like fluctuations or acute changes.

- *Group level (between-effect):* What is the impact of the group's general environment or overall average, which affects all members equally? This captures trait-like, baseline differences or overarching environments.

The difference between these two effects is called the *context effect*. A context effect describes the additional influence that the general environment or baseline trait has on an individual, holding their raw, current state constant. It demonstrates that people with identical current, raw values (such as the exact same momentary income or symptom severity) face different outcomes depending on their baseline or the environment in which they live.

To test whether the within- and between-effects are significantly different from each other, we can estimate their contrast:

```{r}
library(modelbased)
estimate_contrasts(mixed, c("phq4_within", "phq4_between")) |> display(format = "tt")
```

The output shows a significant contrast of 2.41 between the within- and between-effects. Since the between-effect in our model (-6.13) is stronger (more negative) than the within-effect (-3.72), the context-effect (Between minus Within) is -2.41.

**What does this mean practically?**

Imagine two patients who both report the exact same raw `phq4` score on a given day. However, Patient A generally suffers from a higher overarching psychological burden (their personal `phq4` average is 1 unit higher than Patient B's). The context effect tells us that, despite experiencing the exact same severity of symptoms today, Patient A's expected Quality of Life is an additional 2.41 points lower than Patient B's simply because of their higher baseline burden. The trait-like baseline carries an extra "penalty" for the quality of life that goes beyond mere day-to-day fluctuations.

# Time trends of within- and between-effects

To investigate whether the impact of `phq4` changes over time, we can extend our model by adding interaction terms between the time of measurement (`time`) and our two centered variables (`phq4_within` and `phq4_between`).

```{r}
mixed <- lmer(
  QoL ~ time * (phq4_within + phq4_between) + education + (1 + time | ID),
  data = qol_cancer
)
model_parameters(mixed, effects = "fixed") |> display(format = "tt")
```

The results table now shows us whether time has a moderating influence on our effects. To see this, we look at the two interaction terms at the bottom of the table:

- Interaction `time × phq4_within`: The coefficient (0.33) is small and not statistically significant (p = 0.592). This means that the negative effect of individual, temporary fluctuations remains largely stable over time. When a patient's psychological burden acutely increases, their quality of life decreases to a very similar extent at any given time point.

- Interaction `time × phq4_between`: This coefficient (-0.66) is also not statistically significant (p = 0.077). The negative sign suggests that the gap between patients potentially widens slightly over time. The "disadvantage" in quality of life experienced by patients with a generally high psychological baseline burden might further increase over the course of the study compared to less burdened patients.

These temporal dynamics can be best understood visually using Estimated Marginal Means. The generated plots confirm the statistical results from the table very clearly.

```{r}
estimate_means(mixed, c("time", "phq4_within=[sd]")) |> plot()
```

In the first plot (*within-effects over time*), the lines for the different levels of `phq4_within` (mean as well as +/- one standard deviation) run almost parallel. The constant distance between the lines visualizes the lack of interaction: An acute increase in psychological symptoms (shifting to the blue line) depresses the quality of life uniformly across all time points.

```{r}
estimate_means(mixed, c("time", "phq4_between=[sd]")) |> plot()
```

In the second plot (*between-effects over time*), a slight divergence of the lines is visible. While patients with a generally low burden (red line) experience a slight increase in their quality of life over time, the quality of life for patients with a generally high burden (blue line) stagnates at a lower level or even drops minimally.

# Time trends of context effects

Finally, we might ask whether the context effect itself - the difference between the within- and between-effects - is stable, or if the "penalty" of a high baseline burden changes over the course of the study.

Before calculating the context effect directly, it is helpful to first look at the marginal effects (slopes) of both the within- and between-components separately at each time point. This helps us understand the underlying dynamics.

```{r}
# average marginal effect of within-effect at each time point
estimate_slopes(mixed, "phq4_within", by = "time") |> display(format = "tt")
# average marginal effect of between-effect at each time point
estimate_slopes(mixed, "phq4_between", by = "time") |> display(format = "tt")
```

Looking at these separate slopes reveals two opposing trends. The acute impact of a temporary symptom spike (`phq4_within`) slightly decreases in magnitude over time (shifting from -4.04 at Time 1 to -3.39 at Time 3). Conversely, the detrimental impact of a chronically high baseline burden (`phq4_between`) becomes progressively more severe, worsening from -5.36 to -6.68.

Because these two effects drift further apart over time, we can now formally test their difference &mdash; the context effect &mdash; by calculating the marginal contrasts at each specific time point.

```{r}
estimate_contrasts(
  mixed,
  c("phq4_within", "phq4_between"),
  by = "time"
) |>
  display(format = "tt")
```

The contrast analysis reveals a clear and interesting trajectory: the context effect grows substantially stronger as time progresses.

- *At Time 1 (Baseline):* The difference between the within- and between-effect is relatively small (1.32) and not statistically significant (p = 0.188). This indicates that at the beginning of the observations, it does not matter much whether a patient's psychological burden is acute (a temporary spike) or chronic (a generally high baseline). The immediate impact on their Quality of Life is very similar.

- *At Time 2 and Time 3:* As the study progresses, the contrast becomes highly significant and the gap widens (increasing to 2.31, then 3.29).

**What does this mean practically?**

Over time, having a chronically high baseline of psychological symptoms (the trait-level burden) becomes increasingly detrimental compared to merely experiencing a temporary, acute spike in symptoms. While patients might be able to buffer or cope with an acute worsening of their mental state similarly well at any point, the cumulative "wear and tear" of a chronically high burden takes an increasing toll on their quality of life as time goes on.

# Testing the Change in the Context Effect Over Time

While the previous table showed the context effect at each specific time point, we also need to formally test whether the change in this effect over time is statistically significant. We can do this by computing pairwise comparisons of the context effect across the different time points.

```{r}
estimate_contrasts(
  mixed,
  c("phq4_within", "phq4_between", "time")
) |>
  display(format = "tt")
```

This pairwise comparison table adds a crucial statistical caveat to our visual and descriptive observations. The Difference column here represents the mathematical change in the size of the context effect between two specific time points (e.g., the context effect grew by 0.99 points from Time 1 to Time 2).

However, looking at the statistics, you will notice that the estimated difference between any two adjacent time points is exactly identical, yielding the exact same standard error and p-value.

This uniformity is a direct mathematical consequence of our model assuming a *strictly linear time trend*. Under a linear assumption, the rate of change (the slope) is constrained to be constant. Therefore, the estimated change in the context effect from Time 1 to Time 2 is forced to be exactly the same as the change from Time 2 to Time 3. These point-to-point comparisons would only vary if we had modeled time non-linearly (for instance, by adding a quadratic term).

Because a linear model yields constant step-by-step changes, testing these identical individual intervals is often less informative. To test whether the context effect meaningfully changes over time overall, it is more appropriate to evaluate the average contrast of the slopes across the entire study period. To do this, we calculate the contrast between the within- and between-effects without stratifying by time.

```{r}
estimate_contrasts(mixed, c("phq4_within", "phq4_between")) |> display(format = "tt")
```

**What does this mean practically?**

The highly significant overall contrast (2.31, p < .001) confirms that a substantial context effect is at play throughout the entire observation period.

In clinical terms, while an acute, temporary spike in psychological symptoms (the within-effect) definitely harms a patient's quality of life, carrying a chronically high baseline burden (the between-effect) takes a significantly heavier toll. A patient who generally suffers from high psychological distress faces an overarching "trait penalty" that consistently depresses their quality of life more severely than mere day-to-day fluctuations would suggest.

# Trends of Context Effects Across Educational Levels

To explore whether socioeconomic factors influence these dynamics, we can test if the context effect varies across different educational backgrounds. A reasonable assumption might be that highly educated individuals possess more resources (cognitive, financial, or social) and thus cope better with psychological burden.

To formally test this, we extend our model to include a three-way interaction between `time`, `education`, and our centered `phq4` variables.

```{r}
mixed <- lmer(
  QoL ~ time * education * (phq4_within + phq4_between) + (1 + time | ID),
  data = qol_cancer
)
estimate_contrasts(
  mixed,
  c("phq4_within", "phq4_between"),
  by = "education"
) |>
  display(format = "tt")
```

The marginal contrasts analysis yields nuanced results that add an important layer to our understanding of the context effect:

- *Low Education:* The contrast (1.69) is not statistically significant (p = 0.192). For these patients, there is no meaningful difference between an acute symptom spike and a chronically high baseline. Both states depress their quality of life similarly.
- *Mid Education:* The contrast (3.92) is large and highly significant (p < .001). This group actually drives the overall context effect we observed in the previous models. For middle-educated patients, a chronically high psychological burden carries a massive additional penalty compared to a temporary acute spike.
- *High Education:* Interestingly, the contrast reverses its sign (-1.76) but is not statistically significant (p = 0.337). This indicates that for highly educated patients, the context effect disappears entirely.

**What does this mean practically?**

The hypothesis that highly educated people "fare better" is supported here, but in a very specific, mechanistic way. While we would need to look at the main effects to see if their *absolute* Quality of Life is higher, this contrast analysis tells us how they *process* psychological burden.

Highly educated patients appear to be buffered against the specific "chronic penalty". For them, carrying a chronic baseline burden is no more destructive to their quality of life than experiencing an acute, temporary spike. They likely possess the resources needed to manage a chronic psychological load without letting it compound into an overarching environmental penalty.

Conversely, the middle-educated group represents a highly vulnerable population regarding these dynamics. They suffer disproportionately from a chronically high baseline burden.

# Are the Trends of Context Effects Significantly Different Between Groups?

In the previous step, we observed that the context effect seemed to be entirely driven by the middle-educated group, while highly educated patients appeared to be buffered against it. However, to rigorously test whether the size of the context effect is statistically different *between* these groups, we need to compute pairwise comparisons across the educational levels.

We can do this by adding the grouping variable (`education`) as a third term to our contrast statement.

```{r}
estimate_contrasts(
  mixed,
  c("phq4_within", "phq4_between", "education")
) |>
  display(format = "tt")
```

The output now displays the mathematical difference in the size of the context effect between two specific groups.

- *Mid vs. Low (p = 0.154) & High vs. Low (p = 0.124):* The differences involving the low-education group are not statistically significant. The confidence intervals are quite wide, suggesting high variance or a smaller sample size within this specific intersection of the data.
- *High vs. Mid (Difference = -5.69, p = 0.005):* This is the crucial finding. The context effect for the highly educated group is significantly smaller (by 5.69 points) than for the middle-educated group.

**What does this mean practically?**

This pairwise comparison formally solidifies our previous suspicion: The buffering effect of higher education is statistically robust.

It proves that the "chronic penalty" for long-term psychological burden does not hit everyone equally. The structural or psychological advantages possessed by the highly educated group (such as better access to support networks, financial stability, or coping resources) create a mathematically significant difference in how chronic distress is processed. Compared directly to the middle-educated tier—who bear the full weight of the context effect—highly educated patients are significantly better protected from the cumulative "wear and tear" of a chronically high baseline burden.

# Disclaimer

Do not base your conclusions solely on the rigid dichotomy of statistical significance.

# References
