---
title: "Understanding marginalization methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding marginalization methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  out.width = "100%",
  dpi = 300,
  message = FALSE,
  warning = FALSE,
  package.startup.message = FALSE
)

options(
  modelbased_join_dots = FALSE,
  modelbased_estimate = "typical",
  modelbased_select = "minimal",
  parameters_select = "{estimate} ({ci})"
)

pkgs <- c(
  "datawizard",
  "parameters",
  "marginaleffects",
  "emmeans"
)

if (!all(sapply(pkgs, requireNamespace, quietly = TRUE))) {
  knitr::opts_chunk$set(eval = FALSE)
}

if (getRversion() < "4.1.0") {
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{css, echo=FALSE, eval = TRUE}
.custom_note {
  border-left: solid 5px hsl(220, 100%, 30%);
  background-color: hsl(220, 100%, 95%);
  padding: 10px;
  margin-bottom: 10px;
  font-size: 0.9em;
}
```

This vignette explores the `estimate_means()` function, providing a practical guide to calculating and interpreting estimated marginal means. Using the *penguins* dataset, we will demonstrate how different estimation methods (triggered via the `estimate` argument) - `"specific"`, `"typical"`, `"average"`, and `"counterfactual"` - change the assumptions behind your model predictions. By manually recreating these calculations alongside the automated functions, you will gain a much clearer understanding of what actually happens under the hood when interpreting (linear) models and their estimated marginal means.

::: custom_note
One crucial distinction is between *conditional* and *marginal* predictions. This becomes even more important for mixed-effects models, [as outlined in this vignette](https://easystats.github.io/modelbased/articles/mixed_models.html).
:::

First, let's load the necessary packages and dataset.

```{r}
library(easystats)
data(penguins)
```

We fit a simple linear model with a continuous outcome (`flipper_len`), a categorical focal predictor (`sex`), and a categorical non-focal predictor (`species`).

```{r}
m <- lm(flipper_len ~ sex + species, data = penguins)
```

Next, we extract the data used to fit the model. This does not perfectly correspond to the original dataset because a few cases with missing values were removed via list-wise deletion. We will need this data later. Let's also print the regression coefficients.

```{r}
model_data <- get_data(m)

mp <- model_parameters(m)
mp
```

## 1. "Specific" Observation

What happens here? Numeric variables are set to their mean, and factors are set to their reference level. Thus, a prediction is made for a very "specific" observation in the dataset (here: `species = Adelie`).

```{r}
estimate_means(m, "sex", estimate = "specific")

# Alternative code (same result, different wrapper)
estimate_relation(m, by = "sex")
```

To illustrate, we can calculate this manually. "Female" is the reference level for `sex` here. Since `species` is kept at its reference level (Adelie), the estimated marginal mean corresponds exactly to the intercept.

```{r}
# female (reference)
mp$Coefficient[1]
```

For "male", we add the main effect for `sexmale` to the intercept. `species` still remains at the reference level (Adelie).

```{r}
# male
mp$Coefficient[1] + mp$Coefficient[2]
```

### Excursus: How does `predict()` work in comparison?

Let's compare this logic to the standard `predict()` function.

```{r}
pred <- predict(m)

# Let's look at the real data of the 1st observation in the dataset:
penguins[1, c("species", "sex")] # Adelie, male

# Predicted value (predicted mean) for this 1st observation:
pred[1]

# Manual calculation: Intercept + effect for "male" (since Adelie is the reference)
mp$Coefficient[1] + mp$Coefficient[2]

# Let's look at the real data of the 221st observation in the dataset:
penguins[221, c("species", "sex")] # Gentoo, female

# Predicted value (predicted mean) for this 221st observation:
pred[221]

# Manual calculation:
# Intercept (female is reference) + effect for species "Gentoo" (4th coefficient)
mp$Coefficient[1] + mp$Coefficient[4]
```

## 2. "Typical" Observation

What happens here? A data grid is created across all combinations of predictors. The predictions are then averaged across the levels of the non-focal terms (here: `species`). Each level of `species` is weighted *equally* in the calculation, regardless of how often it occurs in the actual empirical data.

```{r}
# We manually create a reference grid with evenly distributed levels
# of our non-focal ("controlled for") term "species".
d <- rbind(
  get_datagrid(m, c("sex", "species = 'Adelie'")),
  get_datagrid(m, c("sex", "species = 'Chinstrap'")),
  get_datagrid(m, c("sex", "species = 'Gentoo'"))
)

# Now we calculate the predicted values for this balanced grid.
# The non-focal terms are weighted exactly equally here
# (1/3 Adelie, 1/3 Chinstrap, 1/3 Gentoo).
d$predicted <- predict(m, newdata = d)

# Manual calculation of the means grouped by "sex":
means_by_group(d, "predicted", "sex")

# This corresponds to the easystats default (estimate = "typical").
estimate_means(m, "sex")
```

## 3. "Average" Observation

What happens here? The prediction is made for *every* single observation in the real dataset. Afterwards, the average of these predictions is calculated per focal group. By doing this, the non-focal terms retain their exact empirical distribution.

```{r}
d <- model_data

# We use the predict() function for the entire original dataset
d$predicted <- predict(m)

# Afterwards, we calculate the mean of the predictions, grouped by "sex"
means_by_group(d, "predicted", "sex")

# The corresponding easystats call:
estimate_means(m, "sex", estimate = "average")
```

## 4. "Counterfactual" Observation (also "Population")

What happens here? To resolve confounding and simulate a pseudo-randomization (G-Computation), we replicate the entire dataset for each level of our focal predictor (`sex`).

We basically force every penguin to be female once, and male once, but we retain its original `species` (and de facto all other unobserved characteristics/confounding that we want to equalize through randomization).

```{r}
# We duplicate ("clone") our dataset to imitate randomization.
d <- do.call(rbind, replicate(2, model_data, simplify = FALSE))

# For each cloned dataset, we set our focal term to one of the two levels.
# The first half of the dataset becomes entirely "female", the second half entirely "male".
d$sex <- as.factor(rep(levels(model_data$sex), each = nrow(model_data)))

# We calculate the predicted values for this new "what-if" scenario,
# i.e., for our pseudo-randomized sample.
d$predicted <- predict(m, newdata = d)

# Now we calculate the average predicted value for the levels of "sex".
# The non-focal terms are weighted exactly proportional to their actual
# occurrence in the real data here.
means_by_group(d, "predicted", "sex")

# The corresponding easystats call:
estimate_means(m, "sex", estimate = "counterfactual")
```

---

## Summary

Broadly speaking, these four methods can be grouped into two conceptual categories: *conditional* predictions and *marginal* predictions. The first two approaches rely on artificially constructed reference grids, providing predictions *conditional* on specific, theoretical individual levels. In contrast, the latter two approaches calculate predictions for the actual empirical observations before averaging, thus reflecting *marginal* predictions at the sample or population level.

### 1. Specific Observation (`estimate = "specific"`)

* **Calculation:** Numeric values are set to their mean. Factors are set to their reference level.
* **Meaning:** The estimated means represent a rather "theoretical" view of the data. A prediction is made for a specific individual from the sample (i.e., a very specific combination of predictor values). This answers the question about the expected average value of the target variable for a *specific* observation.

### 2. Typical Observation (`estimate = "typical"`, the default)

* **Calculation:** Predictions are made for observations represented by a data grid. This grid is formed from all combinations of the predictor levels of the focal terms. For non-focal numeric predictors, the mean is taken, and we marginalize over the factor levels of the non-focal terms (a kind of weighted average is calculated).
* **Meaning:** These predictions are useful to compare defined "groups" and represent the sample well. It answers the question: *"What would the average outcome be for a 'typical' observation?"*, where 'typical' refers to subjects represented by the balanced data grid.

### 3. Average Observation (`estimate = "average"`)

* **Calculation:** Predictions are made for every observation in the sample. Afterwards, the average of all predictions within all groups (or levels) of the focal terms defined in `by` is calculated.
* **Meaning:** These predictions are the closest representation of the actual sample, as the average is taken over the entire dataset. The groups are not represented by a balanced data grid here, but by the empirical distributions of the sample characteristics. It answers the question about the predicted value for an average observation from a specific group within your own data.

### 4. Counterfactual Observation (`estimate = "population"` / `"counterfactual"`)

* **Calculation:** Non-focal predictors are marginalized over the observations of the sample, with the sample being replicated multiple times to create "counterfactuals" (what-if scenarios). Then, the average of these predicted values is calculated (aggregated/grouped by the focal terms).
* **Meaning:** This can be viewed as an extrapolation to a hypothetical target population. Counterfactual predictions are useful because the results can also be transferred to other contexts (G-Computation, causal inference). It answers the question: *"What is the predicted response value for the average observation in the broader target population?"* (or for my "pseudo-randomized" sample). Due to the pseudo-randomization approach, we're allowed to make causal inferences even for observational data (note that there are certain assumptions that need to be fulfilled for causal inference; see section **Assumptions for causal identification** in [this vignette](https://easystats.github.io/modelbased/articles/practical_causality.html)).

### Conclusion: Conditional vs. Marginal Predictions

When deciding which method to use, the core question is whether your estimand relates to a theoretical individual or the population as a whole. Use *conditional* predictions (`"specific"`, `"typical"`) when you want to evaluate expected outcomes for predefined, fixed combinations of covariates (using a reference grid). Use *marginal* predictions (`"average"`, `"counterfactual"`) when you want to account for the actual empirical distribution of your sample. Marginal methods average over the real-world variability of your subjects, making them the recommended choice for making population-level generalizations or drawing causal inferences.

::: custom_note
<p>For **visualizations**, *conditional* predictions or `estimate = "population"` typically yield smoother curves for continuous focal predictors, whereas *marginal* predictions with `estimate = "average"` may introduce noisy visual artifacts. `estimate = "average"` calculates the average based only on the data points that actually exist and doesn't generate a *complete* grid of all theoretical combinations of predictor values. Consequently, the output may not include all the values.</p>
:::

## Appendix: Code equivalent in different R packages

While this vignette focuses on the **easystats** ecosystem, it is helpful to see how these concepts translate to other popular packages. Below, we provide the equivalent R code to calculate the `"specific"`, `"typical"`, `"average"`, and `"counterfactual"` marginal means using the *marginaleffects* and *emmeans* packages. Notice how different packages use different terminology - such as varying `newdata` arguments or specific weighting schemes - to achieve the exact same conceptual estimates under the hood.

We use the `model_parameters()` function to ensure the output displays enough decimal places for a precise comparison.

### Specific Observation

```{r}
# easystats: Sets numeric variables to their mean and factors
# to their reference level.
estimate_means(m, "sex", estimate = "specific")

# marginaleffects: The 'newdata = "mean"' argument replicates this
# "theoretical" observation behavior.
marginaleffects::avg_predictions(m, by = "sex", newdata = "mean") |> model_parameters()
```

### Typical Observation

```{r}
# easystats: Builds a balanced reference grid and weights each group
# equally (the default).
estimate_means(m, "sex", estimate = "typical")

# marginaleffects: 'newdata = "grid"' creates a similar balanced
# data grid.
marginaleffects::avg_predictions(m, by = "sex", newdata = "grid") |> model_parameters()

# emmeans: Equal weighting across a reference grid is the default
# behavior in emmeans.
emmeans::emmeans(m, "sex") |> model_parameters()
```

### Average Observation

```{r}
# easystats: Calculates predictions for every observation in the
# empirical data and averages them.
estimate_means(m, "sex", estimate = "average")

# marginaleffects: Using 'by' without modifying 'newdata' averages
# predictions over the original empirical dataset.
marginaleffects::avg_predictions(m, by = "sex") |> model_parameters()

# emmeans: 'weights = "cell"' weights the reference grid based on
# actual cell frequencies in the data.
emmeans::emmeans(m, "sex", weights = "cell") |> model_parameters()
```

### Counterfactual / Population Observation

```{r}
# easystats: Uses G-computation (cloning the dataset to
# simulate pseudo-randomization).
estimate_means(m, "sex", estimate = "population")

# marginaleffects: Using the 'variables' argument automatically
# triggers this counterfactual approach.
marginaleffects::avg_predictions(m, variables = "sex") |> model_parameters()

# emmeans: 'weights = "proportional"' weights the means
# proportionally to the marginal frequencies of the sample.
emmeans::emmeans(m, "sex", weights = "proportional") |> model_parameters()
```
